---
title: "Multivariate imputation with mice package in R"
author: "Ngoc Tran Trung"
date: "19 December 2016"
layout: post_rmd
comments: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is a research experiment of using `mice` package (Buuren and Groothuis-Oudshoorn, 2011) to impute missing values in multivariate data. Basically, `mice` uses __Fully Conditional Specification (FCS)__ method, whereas _chained equation_ is considered an alias of that term. According to Liu and De (2015), FCS is more relaxed than the conventional method Joint Modeling (JM) in terms of specifiying distribution of the missing data.

In this experiment, we will use an open source dataset __Titanic__ (Kaggle, 2012) as the sample data of the experiment.


#### 1. Multiple imputation vs Single imputation
Both multiple and single imputation techniques are ones that can be used to replace missing values in the data by plausible values. In single imputation, we substitute the missing values of the data by __a single value__. For example, we can replace missing values of a variable by the most common value, or the _mean/median_ value (in case that variable is continuous). Multiple imputation is different in a sense that missing values are replaced by __multiple plausible values__. Using `mice` package, we can even have multiple version of imputed data by trying to impute the data more than one time.


#### 2. Notation

- $\mathbf{Y_j}$ $(j = 1,..., p)$: one of $p$ incomplete variables.
- The observed and missing parts of $Y_j$ are denoted by $\mathbf{Y_j^{obs}}$ and $\mathbf{Y_j^{mis}}$.
    - $Y^{obs} = (Y^{obs},...,Y^{obs}_p)$: observed data in Y.
    - $Y^{mis} = (Y^{mis},...,Y^{mis}_p)$: missing data in Y.
- $\mathbf{m \ge 1}$: the number of imputation.
- $\mathbf{Q}$: predicted model that would be trained on $Y$.

#### 3. Features
`mice` introduces a _modular approach_ to impute data. The imputation framework offered by `mice` comprises of 3 steps: _imputation_, _analysis_ and _pooling_ (Figure 1).

```{r fig.width=8, fig.height=8,echo=FALSE}
library(png)
library(grid)
img <- readPNG("figs/modular_flow.png")
grid.raster(img)
```

- `mice()` Impute incomplete data, creating imputed versions of the data $Y^{(1)}$, $Y^{(2)}$, ..., $Y^{(m)}$.
- `with()` Estimate the predicted model $Q$ on each imputed dataset $Y^{(i)}$, creating multiple versions of the model, which are $\hat{Q}^{(1)}$, $\hat{Q}^{(2)}$,..., $\hat{Q}^{(m)}$
- `pool()`   Pool $\hat{Q}^{(1)}$, $\hat{Q}^{(2)}$,..., $\hat{Q}^{(m)}$ into one estimates $\bar{Q}$

A notable feature of this imputation model is that we use our predicted model of interest $Q$ to analyse the imputed data by ourself. By doing that, we can specifically select the best imputed data that suits our investigation.

#### 4. Experiment

#### 4.1. Missing value inspection
First and foremost, load the package and data. In this experiment, we will use Titanic data free available from Kaggle.

```{r, message=FALSE}
# Load library
library(mice) # Imputation
library(VIM) #aggr, marginplot
library(lattice) #stripplot
# Load data
dt <- read.csv("data/titanic.csv", na.strings = c("NA", ""))
# Inspect the data structure
str(dt, strict.width = "wrap") 

# Remove unnecessary columns
dt[, c("PassengerId", "Name", "Cabin", "Ticket")] <- NULL
# Summary the data
summary(dt) 

# Fare = 0 doesn't make sense, set 0 values in Fare as missing values
dt$Fare[dt$Fare == 0] <- NA
```

Our data contains a lot of missing values. We will use different methods to inspect them. First, we will use `md.pattern` method to inspect missing-data patterns.

```{r}
md.pattern(dt)
```

Let's intepret information from the retrieved table. Note that in the table, 0 represents missing in the data. Look at the table by row, the table tells us that we have 705 complete rows, 169 rows where only `Age` is missing, `7` rows where only `Fare` is missing, 2 rows where `Embarked` is missing, and 8 rows where both `Fare` and `Age` are missing. Vertically, the tables tells us that `Embarked` has 2 missing values, `Fare` 15, and `Age` 177. In overall, we have 194 missing values in the data, and 169 + 7 + 2 + 8 = 186 incomplete rows.

Besides `md.pattern`, we can visually inspect missing patterns of the data using the package `VIM`. First, we will use `aggr` to aggregate missing values. 

```{r, message=FALSE}
aggr(dt, col = c("navyblue","red"), sortVars = TRUE, numbers = TRUE)
```

The method `aggr` is essentially similar to `md.patterns`, except that `aggr` displays missing values in portion. For example, roughly 20% (177 rows) of `Age` are incomplete, 1.6% (15 rows) of `Fare` missing, and 0.2% (2 rows) of `Embarked` missing. There are roughly 0.9% (8 rows) of the data where both `Age` and `Fare` are missing. Thus, it seems more convenient when combine both `md.pattern` and `aggr` together to inspect missing values of the data.

We can also observe missing values by pair. The function `marginplot` is perfect for this purpose. We will use `marginplot` to inspect missing patterns of the pair `Age` and `Fare`.

```{r}
marginplot(dt[, c("Fare", "Age")], col = c("navyblue", "red"), 
           numbers = TRUE, pch = 19)
```

Let's intepret information given by the plot. 

- Blue points in the main area indicates the number of rows that both `Age` and `Fare` are complete. The red points in the left side indicates missing values of `Fare` when `Age` is observed. Similarly, the red points in the bottom margin indicates missing values of `Age` when `Fare` is observed. 
- Three numbers tells us that there are 15 missing values in `Fare` when `Age` is observed, 177 missing values in `Age` when `Fare` is observed, and 8 commonly missing values for both.
- The blue and red boxplots summarise the marginal distribution of complete and incomplete rows. For example, the distribution of complete `Age` values is (0, 80) years old, the distribution of missing value in `Fare` when `Age` is observed is from (20, 50) years old (i.e., 15 passengers who has missing values of `Fare` are from 20 to 50 years old).

##### 4.2. Data imputation with `mice()`
We can simply impute the data by using the function `mice()`. For example,

```{r, message=FALSE, results = "hide"}
# Impute missing values
imp <- mice(dt, maxit = 20) 
```

By default, `mice()` will create __5 version of imputed data__ (i.e., `m = 5`), and for each version, the default number of iterations is also 5 (i.e., `maxit = 5`). According to the authors, the convergence of Gibbs sampling often happens when `maxit` is between 10 and 20. 

```{r}
# Extract the 1st version of the imputed data using complete()
dt_imp_1 <- complete(imp, 1)
summary(dt_imp_1)
```
We can see that missing values no longer exist in the data. Let's see how the data is imputed across 5 versions. First begin with `Embarked`

```{r}
# Embarked
idx <- which(is.na(dt$Embarked)) # get index of missing values in the original data
sapply(c(1:5), function(x) {(complete(imp, x)$Embarked[idx])})
```

So, `mice()` imputed missing values in `Embarked` with different values. Note that in the original data, there are 644 rows of `S`. So I suspect the imputed values are likely (S,S).

Next, we will explore the distribution of imputed values of `Age` and `Fare`. It can be seen from both plots that the distribution of imputed values are similar to that of observed values. 

```{r}
# Age
dt_imp_1 <- complete(imp, "long", inc = TRUE)
col <- rep(c("blue", "red")[1+as.numeric(is.na(imp$data$Age))], 6)
stripplot(Age~.imp, data = dt_imp_1, jit = TRUE, fac = 0.8, col = col, 
          xlab = "Imputation Number")

# Fare
col <- rep(c("blue", "red")[1+as.numeric(is.na(imp$data$Fare))], 6)
stripplot(Fare~.imp, data = dt_imp_1, jit = TRUE, fac = 0.8, col = col, 
          xlab = "Imputation Number")
```

##### 4.2. Data analysis with `with()` 
After imputation, we will analyse imputed data using `with()`. As said, we analyse the imputed data by our predict model of interest $Q$.

```{r}
fit <- with(imp, glm(Survived ~ Sex + Age + Fare + Pclass + SibSp + Parch 
                     + Embarked, family = "binomial"))
```

##### 4.3. Result pooling with `pool()`

```{r}
pool(fit)
summary(pool(fit))
```

#### 5. Convergence assessment
Convergence assessment of Gibbs sampling algorithm is important to check that if the algorithm is converged. According to the paper: _"There is no clear-cut method for determining whether the Gibbs sampling algorithm has converged. What is often done is to plot one or more parameters against the iteration number"_ and _"on convergence, the different streams should be freely intermingled with each other, without showing any definite trends. Convergence is diagnosed when the variance between different sequences is no larger than the variance with each individual sequence"_.

```{r}
plot(imp)
```

#### 6. Conclusion
In this experiment, we have demonstrated how missing values in the data can be imputed using FCS technique. When compare FCS with JM technique, MICE is considered as a more adaptable method when we can select appropriate imputation models for  variables on a case by case basis. For example, we can choose linear regression or Bayesian linear regression for continuous variables, and logistic regression or linear discriminant analysis for nominal variables.

`mice` package in R is a powerful and convenient library that enables multivariate imputation in a modular approach consisting of three subsequent steps. First, we can impute missing values by using a single `mice()` function, then effectively analyse imputed versions of data by using `with()` method with our own model of choice, and finally report the imputation result by using `pool()` method. 

After this experiment, we believe that `mice` package is capable of supporting our future research experiments, where we would have chance to explore additional features of it, such as, implementing our own imputation models.

#### References
Buuren, S. & Groothuis-Oudshoorn, K. (2011, December). MICE: Mulitivariate Imputation by Chained Equations in R. Journal of Statistical Software, 45(3), 1–67. Retrieved from http://doc.utwente.nl/78938/

Kaggle. (2012, September). Titanic: machine learning from disaster. Retrieved from https://www.kaggle.com/c/titanic

Liu, Y. & De, A. (2015, July). Multiple Imputation by Fully Conditional Specification for Dealing with Missing Data in a Large Epidemiologic Study. International Journal of Statistics in Medical Research, 4(3), 287–295. doi:10.6000/1929-6029.2015.04.03.7